---
title: "Appendix A2. Regression Discontinuity (Sharp RD) Example"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    number-sections: false
execute:
  warning: false
  message: false
  cache: true
  freeze: auto
---

## Overview

This appendix walks through a simple *sharp* regression discontinuity design (RDD) using simulated data.\
The running variable is `x` with a cutoff at `50`, treatment is assigned deterministically (`D = 1[x ≥ 50]`), and the outcome `y` has a true jump at the cutoff.

We follow a basic RDD workflow: 1) simulate data,\
2) pick a bandwidth,\
3) estimate the discontinuity (with and without covariates),\
4) run a few standard checks (density/manipulation; covariate balance),\
5) do a quick bandwidth sensitivity check, and\
6) run [konfound](https://cran.r-project.org/web/packages/konfound/index.html) on the main estimate.



## A2.1 Setup

```{r}
#| label: a2-setup
#| include: true

set.seed(5)

# General data and plotting
library(tidyverse)

# RDD estimation
library(rdrobust)
library(rddensity)  # Cattaneo et al. density test + plot

# Sensitivity analysis
library(konfound)

# Custom plotting helper used later (expects file to be in the working directory)
source("rdplot_function.R")
```

## A2.2 Simulate a sharp RD dataset

A few small implementation choices here are just to keep things tidy:

-   We cap the running variable to `[0, 100]` to avoid extreme tails.
-   Covariates are generated *independently* of treatment assignment (so balance should hold around the cutoff).

```{r}
#| label: a2-simulate-data
#| include: true

n        <- 5000
cutoff   <- 50
true_tau <- 40  # true local treatment effect (jump at the cutoff)

dat <- tibble(
  x = rnorm(n, 50, 25)
) %>%
  mutate(
    x = if_else(x < 0, 0, x),
    x = if_else(x > 100, 100, x),
    D = if_else(x >= cutoff, 1, 0),  # sharp assignment
    ses          = rnorm(n(), 0, 1),
    female       = rbinom(n(), 1, 0.5),
    sci_interest = rnorm(n(), 0, 1)
  ) %>%
  mutate(
    y = 25 +
      true_tau * D +     # discontinuity
      1.5 * x +          # smooth relationship with x
      5   * ses +
      3   * female +
      4   * sci_interest +
      rnorm(n(), 0, 20)
  )

glimpse(dat)
```

## A2.3 Quick plot (does the jump look plausible?)

This is not “proof” of anything, but it’s a good first check.

If the plot looks flat or noisy right at the cutoff, it’s worth slowing down before trusting estimates.

```{r}
#| label: a2-plot-raw
#| fig-cap: "Raw scatter plot with a vertical line at the cutoff (50)."
#| include: true

ggplot(dat, aes(x, y, colour = factor(D))) +
  geom_point(alpha = 0.3) +
  geom_vline(xintercept = cutoff, colour = "grey40", linetype = 2) +
  stat_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Running variable (X)",
    y = "Outcome (Y)",
    colour = "Treatment (D)"
  )
```

## A2.4 Bandwidth selection (triangular kernel)

Here we use `rdbwselect()` and take the main MSE-optimal bandwidth.

This bandwidth is the default starting point for the main estimate.

```{r}
#| label: a2-bandwidth
#| include: true

bw_tri <- rdbwselect(
  y      = dat$y,
  x      = dat$x,
  c      = cutoff,
  kernel = "triangular"
)

# Main MSE-optimal bandwidth (first row/column in the bws matrix)
main_bw <- bw_tri$bws[1, 1]
main_bw
```

## A2.5 Main RD estimate (with and without covariates)

Two versions:

-   **Simple RD:** outcome on running variable only.
-   **Covariate-adjusted RD:** adds covariates as precision controls (should not change identification in sharp RD).

```{r}
#| label: a2-rd-main-simple
#| include: true

rd_main_simple <- rdrobust(
  y      = dat$y,
  x      = dat$x,
  c      = cutoff,
  h      = main_bw,
  kernel = "triangular"
)

summary(rd_main_simple)
```

```{r}
#| label: a2-rd-main-cov
#| include: true

rd_main_cov <- rdrobust(
  y      = dat$y,
  x      = dat$x,
  c      = cutoff,
  h      = main_bw,
  kernel = "triangular",
  covs   = cbind(dat$ses, dat$female, dat$sci_interest)
)

summary(rd_main_cov)
```

## A2.6 Manipulation / density checks around the cutoff

A common RDD worry is that people can “sort” around the cutoff (i.e., manipulate the running variable).

Density discontinuity tests are one way to flag that kind of problem.



-   Cattaneo et al. density test (rddensity)

```{r}
#| label: a2-rddensity
#| include: true

dens <- rddensity(
  X          = dat$x,
  c          = cutoff,
  h          = main_bw,
  vce        = "jackknife",
  massPoints = FALSE
)

summary(dens)
```

```{r}
#| label: a2-rdplotdensity
#| fig-cap: "Density plot around the cutoff using rdplotdensity()."
#| include: true

rdplotdensity(
  dens,
  dat$x,
  xlabel = "Running Variable",
  ylabel = "Density"
)
```

## A2.7 Covariate balance near the cutoff

If RDD is behaving as expected, covariates should be smooth at the cutoff.

This isn’t a formal proof, but obvious jumps in pre-treatment covariates are a red flag.

```{r}
#| label: a2-center-running-var
#| include: true

dat1 <- dat %>%
  mutate(x_centered = x - cutoff)
```

```{r}
#| label: a2-cov-balance-female
#| fig-cap: "Covariate balance check at the cutoff: female."
#| include: true

plot.RD(
  female ~ x_centered,
  treat  = "D",
  data   = dat1,
  yaxt   = "n",
  linecol = "black",
  kdens  = "rectangular",
  h      = main_bw,
  cex    = 0.5,
  main   = "Covariate Balance Check: female",
  ylab   = "female",
  xlab   = "Running Variable (Centered)",
  xlim   = c(-50, 50)
)
axis(side = 2, at = c(0, 1), labels = c(0, 1))
```

```{r}
#| label: a2-cov-balance-ses
#| fig-cap: "Covariate balance check at the cutoff: SES."
#| include: true

plot.RD(
  ses ~ x_centered,
  treat  = "D",
  data   = dat1,
  linecol = "black",
  kdens  = "rectangular",
  h      = main_bw,
  cex    = 0.5,
  main   = "Covariate Balance Check: SES",
  ylab   = "Socioeconomic Status (SES)",
  xlab   = "Running Variable (Centered)",
  xlim   = c(-50, 50)
)
```

```{r}
#| label: a2-cov-balance-interest
#| fig-cap: "Covariate balance check at the cutoff: science interest."
#| include: true

plot.RD(
  sci_interest ~ x_centered,
  treat  = "D",
  data   = dat1,
  linecol = "black",
  kdens  = "rectangular",
  h      = main_bw,
  cex    = 0.5,
  main   = "Covariate Balance Check: science interest",
  ylab   = "science interest",
  xlab   = "Running Variable (Centered)",
  xlim   = c(-50, 50)
)
```

## A2.8 Bandwidth sensitivity (narrower vs wider)

This is a quick “does the estimate flip if I nudge bandwidth?” check.

It’s not the only robustness check, but it’s an easy one that reviewers often expect.

```{r}
#| label: a2-bandwidth-sensitivity
#| include: true

bw_narrow <- main_bw * 0.7
bw_wide   <- main_bw * 1.3

bw_narrow
bw_wide
```

```{r}
#| label: a2-rd-narrow
#| include: true

rd_narrow <- rdrobust(
  y      = dat$y,
  x      = dat$x,
  c      = cutoff,
  h      = bw_narrow,
  kernel = "triangular"
)

summary(rd_narrow)
```

```{r}
#| label: a2-rd-wide
#| include: true

rd_wide <- rdrobust(
  y      = dat$y,
  x      = dat$x,
  c      = cutoff,
  h      = bw_wide,
  kernel = "triangular"
)

summary(rd_wide)
```

## A2.9 Sensitivity analysis (Konfound)

Here we run `konfound` on the estimate/SE you provided.

If you want this to be fully automatic, we can extract the estimate and SE from `rd_main_simple` instead of hard-coding them.

```{r}
#| label: a2-konfound
#| include: true

pkonfound(
  est_eff      = 38.943,
  std_err      = 11.011,
  n_obs        = 1788,
  n_covariates = 1,
  index        = "IT"
)

pkonfound(
  est_eff      = 38.943,
  std_err      = 11.011,
  n_obs        = 1788,
  n_covariates = 1,
  index        = "RIR"
)
```
