---
title: "Appendix A1. Propensity Score Weighting (CBPS-IPTW)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    number-sections: false
execute:
  warning: false
  message: false
  cache: true
  freeze: auto
---

## Overview

This appendix example estimates the average treatment effect (ATE) of having an IEP on science self-efficacy (`X1SCIEFF`). We (1) build an analytic dataset, (2) estimate propensity score weights using CBPS, (3) check covariate balance, (4) fit a weighted outcome model, and (5) run a sensitivity analysis with [konfound](https://cran.r-project.org/web/packages/konfound/index.html).

## A1.1 Setup

```{r}
#| label: a1-setup
#| include: true

set.seed(12345)

# Packages used in this example
library(dplyr)
library(CBPS)
library(twang)
library(survey)
library(konfound)
```

## A1.2 Load data

```{r}
#| label: a1-load-data
#| include: true
load("hsls_17_student_pets_sr_v1_0.RData")
hsls <- hsls_17_student_pets_sr_v1_0

# Quick sanity check
#dim(hsls)
```

## A1.3 Keep variables used in the example

```{r}
#| label: a1-select-vars
#| include: true

vars_needed <- dplyr::select(hsls,
    STU_ID,        # student ID

    # treatment
    X1IEPFLAG,     # IEP status

    # outcome
    X1SCIEFF,      # science self-efficacy

    # covariates
    X1SEX,         # gender
    X1RACE,        # race
    X1DUALLANG,    # language background
    X1LOCALE,      # locale

    X1SESQ5,       # SES quintile
    P1MARSTAT,     # parents' marital status
    P1HIDEG1,      # parent 1 highest degree
    P1HIDEG2,      # parent 2 highest degree
    P1SLD,         # parent report: learning disability
    P1DD           # parent report: developmental delay
  )

summary(vars_needed)
```

## A1.4 Recode missing categories as NA

Some HSLS variables code missingness as labeled categories.

Here we convert those labels to `NA` so that downstream modeling behaves as expected.

```{r}
#| label: a1-handle-missing-labels
#| include: true

missing_levels <- c(
  "Item legitimate skip/NA",
  "Unit non-response",
  "Missing"
)

vars_clean <- vars_needed %>%
  mutate(across(
    .cols = c(X1IEPFLAG, X1SEX, X1RACE, X1DUALLANG, X1LOCALE,
              X1SESQ5, P1MARSTAT, P1HIDEG1, P1HIDEG2, P1SLD, P1DD),
    .fns = ~ ifelse(as.character(.x) %in% missing_levels, NA, as.character(.x))
  ))

summary(vars_clean)
```

## A1.5 Construct analysis variables (treatment + covariates)

The goal here is not to be clever—just to turn a handful of key survey fields into clean indicators that are easy to use in the weighting model.

```{r}
#| label: a1-construct-variables
#| include: true

analytic <- vars_clean %>%
  mutate(
    # treatment: IEP
    IEP = if_else(X1IEPFLAG == "Student has an IEP", 1, 0, missing = NA),

    # demographics
    female = if_else(X1SEX == "Female", 1, 0, missing = NA),

    race_white = if_else(X1RACE == "White, non-Hispanic", 1, 0, missing = NA),
    race_black = if_else(X1RACE == "Black/African-American, non-Hispanic", 1, 0, missing = NA),
    race_asian = if_else(X1RACE == " Asian, non-Hispanic ", 1, 0, missing = NA),
    race_hispanic = if_else(X1RACE == " Hispanic, race specified ", 1, 0, missing = NA),

    # language / locale
    dual_lang = if_else(X1DUALLANG == "First language is English only", 1, 0, missing = NA),
    locale_rural = if_else(X1LOCALE == "Rural", 1, 0, missing = NA),

    # family background
    divorced = if_else(P1MARSTAT %in% c("Divorced", "Separated"), 1, 0, missing = NA),
    bachelorabove1 = if_else(P1HIDEG1 %in% c("Bachelor's degree", "Master's degree"), 1, 0, missing = NA),
    bachelorabove2 = if_else(P1HIDEG2 %in% c("Bachelor's degree", "Master's degree"), 1, 0, missing = NA),

    # parent-reported conditions
    LD = if_else(P1SLD == "Yes", 1, 0, missing = NA),
    DD = if_else(P1DD == "Yes", 1, 0, missing = NA)
  ) %>%
  mutate(
    # SES: convert ordered quintiles into a simple numeric scale 1–5
    SES = factor(
      X1SESQ5,
      levels = c(
        "First quintile (lowest)",
        "Second quintile",
        "Third quintile",
        "Fourth quintile",
        "Fifth quintile (highest)"
      ),
      labels = c(1, 2, 3, 4, 5)
    ) %>% as.numeric()
  )

summary(analytic)
```

## A1.6 Complete-case analytic sample

For this appendix example we use complete cases.

(If you later decide to use multiple imputation, it can slot in here without changing the rest of the workflow.)

```{r}
#| label: a1-complete-cases
#| include: true

analytic_clean <- analytic[complete.cases(analytic), ]
analytic_clean$IEP <- as.factor(analytic_clean$IEP)

dim(analytic_clean)
table(analytic_clean$IEP)
```

## A1.7 Estimate CBPS weights (ATE)

We use **CBPS** to estimate weights that balance covariates by construction.

Then we normalize the weights so the average weight is 1 (helps with interpretability and avoids tiny/huge totals).

```{r}
#| label: a1-cbps-iptw
#| include: true

cbps_fit <- CBPS(
  IEP ~ female + race_white + race_black +
    dual_lang + locale_rural + SES + divorced +
    bachelorabove1 + bachelorabove2 + LD + DD,
  data = analytic_clean,
  method = "exact",
  ATT = 0
)

analytic_clean$IPTW <- cbps_fit$weights
summary(analytic_clean$IPTW)

# Normalize weights: mean(weight) = 1
analytic_clean$finalWeightATE <- with(analytic_clean, IPTW / mean(IPTW))

with(analytic_clean, by(finalWeightATE, IEP, summary))
```

## A1.8 Covariate balance (before vs after weighting)

This is the gut-check step: if weighting doesn’t improve balance, we don’t trust the ATE estimate.

```{r}
#| label: a1-balance-check
#| include: true

covnames <- c(
  "female", "race_white", "race_black",
  "dual_lang", "locale_rural", "SES", "divorced",
  "bachelorabove1", "bachelorabove2",
  "LD", "DD"
)

balance_before <- bal.stat(
  data = analytic_clean,
  vars = covnames,
  treat.var = "IEP",
  w.all = 1,
  get.ks = FALSE,
  sampw = 1,
  estimand = "ATE",
  multinom = FALSE
)

balance_after <- bal.stat(
  data = analytic_clean,
  vars = covnames,
  treat.var = "IEP",
  w.all = analytic_clean[, "finalWeightATE"],
  get.ks = FALSE,
  sampw = 1,
  estimand = "ATE",
  multinom = FALSE
)

balance_before
balance_after
```

## A1.9 Weighted outcome model (survey regression)

We fit a simple weighted regression of `X1SCIEFF` on `IEP`.

The design object uses the CBPS weights.

```{r}
#| label: a1-weighted-model
#| include: true

design_IPTW <- svydesign(
  id = ~1,
  weights = ~finalWeightATE,
  data = analytic_clean
)

model_IPTW <- svyglm(X1SCIEFF ~ IEP, design = design_IPTW)
summary(model_IPTW)
```

## A1.10 Sensitivity analysis (Konfound)

Here we run `konfound` using the reported estimate and standard error.

In practice, you’ll usually pull the estimate/SE directly from the fitted model output, but the hard-coded values are fine for a fixed appendix example.

```{r}
#| label: a1-konfound
#| include: true

# RIR index
pkonfound(
  est_eff = -0.63613,
  std_err = 0.16460,
  n_obs = 5282,
  n_covariates = 1,
  index = "RIR"
)

# IT index
pkonfound(
  est_eff = -0.63613,
  std_err = 0.16460,
  n_obs = 5282,
  n_covariates = 1,
  index = "IT"
)
```
