---
title: "Appendix A3. Difference-in-Differences (DiD) Example"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    number-sections: false
execute:
  warning: false
  message: false
  cache: true
  freeze: auto
---

## Overview

This appendix illustrates a simple two-period DiD setup using HSLS variables.
The outcome is science self-efficacy measured at baseline (`X1SCIEFF`) and follow-up (`X2SCIEFF`).
Schools are treated if they adopt a PD requirement by follow-up (while having no requirement at baseline).
We estimate a baseline DiD and then a version that adds student covariates for precision.


## A3.1 Setup

```{r}
#| label: a3-setup
#| include: true

set.seed(123)

library(dplyr)
library(tidyr)
library(ggplot2)
library(fixest)
library(konfound)
```

## A3.2 Load data
```{r}
#| label: a3-load-data
#| include: true

load("hsls_17_student_pets_sr_v1_0.RData")
hsls <- hsls_17_student_pets_sr_v1_0

dim(hsls)
```
## A3.3 Select variables used in the example
```{r}
#| label: a3-select-vars
#| include: true

vars_needed <- hsls[, c(
  "STU_ID",
  "SCH_ID",
  "A1MSPDLEARN",
  "A2MSPDLEARN",
  "X1SCIEFF",
  "X2SCIEFF",
  "X1SEX",
  "X1RACE",
  "X1DUALLANG",
  "X1LOCALE",
  "X1SESQ5"
)]

summary(vars_needed)
```
## A3.4 Recode labeled missing categories as NA

HSLS includes several labeled categories that essentially mean "missing" or "not applicable".

For this example, we drop them by recoding to `NA`.

```{r}
#| label: a3-handle-missing
#| include: true

missing_levels <- c(
  "Item legitimate skip/NA",
  "Unit non-response",
  "Missing",
  "Item not administered: abbreviated interview",
  "Component not applicable"
)

vars_clean <- vars_needed %>%
  mutate(across(
    .cols = c(A1MSPDLEARN, A2MSPDLEARN,
              X1SEX, X1RACE, X1DUALLANG, X1LOCALE, X1SESQ5),
    .fns = ~ ifelse(as.character(.x) %in% missing_levels, NA, as.character(.x))
  ))

summary(vars_clean)
```

## A3.5 Construct treatment indicators and covariates

Here `a1` and `a2` indicate whether a PD requirement is present at baseline vs follow-up.

We also create a compact set of student-level covariates.
```{r}
#| label: a3-construct-vars
#| include: true

analytic <- vars_clean %>%
  mutate(
    a1 = if_else(A1MSPDLEARN == "Yes", 1, 0, missing = NA),
    a2 = if_else(A2MSPDLEARN == "Yes", 1, 0, missing = NA),

    female = if_else(X1SEX == "Female", 1, 0, missing = NA),

    race_white = if_else(X1RACE == "White, non-Hispanic", 1, 0, missing = NA),
    race_black = if_else(X1RACE == "Black/African-American, non-Hispanic", 1, 0, missing = NA),
    race_asian = if_else(X1RACE == " Asian, non-Hispanic ", 1, 0, missing = NA),

    dual_lang = if_else(X1DUALLANG == "First language is English only", 1, 0, missing = NA),
    locale_rural = if_else(X1LOCALE == "Rural", 1, 0, missing = NA)
  ) %>%
  mutate(
    SES = factor(
      X1SESQ5,
      levels = c(
        "First quintile (lowest)",
        "Second quintile",
        "Third quintile",
        "Fourth quintile",
        "Fifth quintile (highest)"
      ),
      labels = c(1, 2, 3, 4, 5)
    ) %>% as.numeric()
  )

summary(analytic)
```
## A3.6 Complete-case sample

For a clean appendix example, we use complete cases.

(If your main paper uses a different missing-data strategy, this is the place to swap it in.)
```{r}
#| label: a3-complete-cases
#| include: true

analytic_clean <- analytic[complete.cases(analytic), ]
dim(analytic_clean)
```

## A3.7 Define treated vs never-treated schools (based on adoption)

We restrict to schools with **no requirement at baseline** (`a1 == 0`).

Then:

- treated schools: adopt by follow-up (`a2 == 1`)
- control schools: never adopt (`a2 == 0`)
```{r}
#| label: a3-define-groups
#| include: true

hsls_did <- analytic_clean %>%
  filter(a1 == 0) %>%
  mutate(
    treat_school  = ifelse(a1 == 0 & a2 == 1, 1, 0),
    never_treated = ifelse(a1 == 0 & a2 == 0, 1, 0)
  ) %>%
  filter(treat_school == 1 | never_treated == 1)

table(hsls_did$treat_school)
```
## A3.8 Simulate school IDs for clustering (SCH_ID_SIM)

In a proper DiD, you normally already have a stable school identifier for clustering.

Here we simulate a school ID assignment to keep the example self-contained (and to show the mechanics of clustering in `fixest`).
```{r}
#| label: a3-simulate-school-ids
#| include: true

# Student counts by adoption status at follow-up
n_treatment_students <- sum(hsls_did$a2 == 1, na.rm = TRUE)
n_control_students   <- sum(hsls_did$a2 == 0, na.rm = TRUE)

# Allocate schools proportionally (tweak if you want)
n_total_schools <- 80
prop_treatment <- n_treatment_students / (n_treatment_students + n_control_students)

n_schools_treatment <- round(n_total_schools * prop_treatment)
n_schools_control   <- n_total_schools - n_schools_treatment

# Assign simulated school IDs
hsls_did$SCH_ID_SIM <- NA_integer_

idx_treatment <- hsls_did$a2 == 1
hsls_did$SCH_ID_SIM[idx_treatment] <-
  sample(1:n_schools_treatment, sum(idx_treatment), replace = TRUE)

idx_control <- hsls_did$a2 == 0
hsls_did$SCH_ID_SIM[idx_control] <-
  sample((n_schools_treatment + 1):(n_schools_treatment + n_schools_control),
         sum(idx_control), replace = TRUE)

table(hsls_did$SCH_ID_SIM)[1:10]
```
## A3.9 Reshape to long format (two rows per student)

We stack baseline and follow-up outcomes, then create:

- `post`: 0 baseline, 1 follow-up
- `treat_post`: DiD interaction term
```{r}
#| label: a3-long-format
#| include: true

cols_to_keep <- c(
  "STU_ID", "SCH_ID_SIM", "treat_school",
  "X1SCIEFF", "X2SCIEFF",
  "female", "race_white", "race_black", "race_asian",
  "dual_lang", "locale_rural", "SES"
)

hsls_long <- hsls_did[, cols_to_keep] %>%
  pivot_longer(
    cols      = c(X1SCIEFF, X2SCIEFF),
    names_to  = "wave",
    values_to = "SCI_SE"
  ) %>%
  mutate(
    post       = ifelse(wave == "X2SCIEFF", 1, 0),
    treat_post = treat_school * post
  )

table(hsls_long$post)
```
## A3.10 DiD estimation (two-way setup with school fixed effects)

### (a) Basic DiD

This is the clean baseline model: outcome on `post` and `treat_post`, with school fixed effects.
```{r}
#| label: a3-did-basic
#| include: true

did_basic <- feols(
  SCI_SE ~ post + treat_post | SCH_ID_SIM,
  data = hsls_long
)

summary(did_basic)
```

### (b) DiD with student covariates (precision controls)

Adding covariates is not required for identification in DiD, but it can help precision and guard against small sample quirks.
```{r}
#| label: a3-did-covariates
#| include: true

did_cov <- feols(
  SCI_SE ~ post + treat_post +
    female + race_white + race_black + race_asian +
    dual_lang + locale_rural + SES | SCH_ID_SIM,
  data = hsls_long
)

summary(did_cov)
```
## A3.11 Mean trends plot (treated vs control)

This plot is a quick visual check of how treated and control groups move from baseline to follow-up.

With only two time points, we can’t really “test” parallel trends, but at least we can see whether the groups are moving in obviously different ways.
```{r}
#| label: a3-trends
#| fig-cap: "Mean science self-efficacy over time by treatment group."
#| include: true

trend_dat <- hsls_long %>%
  group_by(treat_school, post) %>%
  summarise(
    mean_SCI_SE = mean(SCI_SE, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  trend_dat,
  aes(
    x = post,
    y = mean_SCI_SE,
    group = factor(treat_school),
    linetype = factor(treat_school),
    shape = factor(treat_school)
  )
) +
  geom_line() +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  scale_x_continuous(breaks = c(0, 1), labels = c("Baseline", "Follow-up")) +
  labs(
    x = "Wave",
    y = "Mean science self-efficacy",
    linetype = "Treated school",
    shape = "Treated school",
    caption = "Dashed line marks the intervention period."
  ) +
  theme_minimal()
```
## A3.12 Sensitivity analysis (Konfound)

Below we run `konfound` using the estimate and SE you provided for the DiD effect.
```{r}
#| label: a3-konfound
#| include: true

pkonfound(
  est_eff      = -0.0346,
  std_err      = 0.1025,
  n_obs        = 6356,
  n_covariates = 2,
  index        = "RIR"
)

pkonfound(
  est_eff      = -0.0346,
  std_err      = 0.1025,
  n_obs        = 6356,
  n_covariates = 2,
  index        = "IT"
)
```

