---
title: "Appendix A4. Instrumental Variables (IV / 2SLS) Example"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    number-sections: false
execute:
  warning: false
  message: false
  cache: true
  freeze: auto
---

## Overview

This appendix gives a small IV example using simulated data.  
We generate a setting where the treatment (`friends`, having friends in the classroom) is confounded by an unobserved factor (`motivation`).  
A plausible instrument (`distance`, minutes to school) affects the treatment but does not directly affect test scores except through treatment.

We then estimate the effect of `friends` on `scores` using 2SLS and run a [konfound](https://cran.r-project.org/web/packages/konfound/index.html) sensitivity analysis on the IV estimate.

---

## A4.1 Setup

```{r}
#| label: a4-setup
#| include: true

set.seed(123)

library(AER)      # provides ivreg()
library(konfound)
library(dplyr)
```

## A4.2 Simulate data

A quick note on the simulation logic:

- `motivation` affects both `friends` and `scores` (this is the confounder).
- `distance` shifts the likelihood of `friends` but does not enter the outcome equation directly.

```{r}
#| label: a4-simulate
#| include: true

n <- 2000

# Instrument: distance to school (minutes)
distance <- rnorm(n, mean = 20, sd = 8)

# Unobserved confounder: student motivation
motivation <- rnorm(n)

# Treatment: having friends in the classroom (binary)
prob_friends <- plogis(-0.5 + 0.08 * distance - 0.5 * motivation)
friends <- rbinom(n, 1, prob_friends)

# Outcome: test scores (friends has a true causal effect of +3)
scores <- 50 + 3 * friends + 5 * motivation + rnorm(n, sd = 5)

dat <- data.frame(scores, friends, distance, motivation)

summary(dat)
```

## A4.3 Why IV is useful here (quick check)

If we regress `scores` on `friends` directly, motivation will bias the estimate because it’s omitted and correlated with `friends`.

```{r}
#| label: a4-ols
#| include: true

ols_model <- lm(scores ~ friends, data = dat)
summary(ols_model)
```



## A4.4 2SLS / IV estimation

We instrument `friends` with `distance`:

[

scores = \beta_0 + \beta_1 friends + \varepsilon,\quad friends \leftarrow distance

]

```{r}
#| label: a4-iv
#| include: true

iv_model <- ivreg(scores ~ friends | distance, data = dat)
summary(iv_model)
```


## A4.5 First-stage strength (diagnostic)

This is a practical check: if the instrument barely moves treatment, IV will be noisy and potentially misleading.

`summary(iv_model, diagnostics = TRUE)` reports common diagnostics (including a weak instrument test).

```{r}
#| label: a4-iv-diagnostics
#| include: true

summary(iv_model, diagnostics = TRUE)
```


## A4.6 Sensitivity analysis (Konfound)

Below we run `konfound` using the estimate and SE you provided.

If you want this fully reproducible end-to-end, we can extract these values directly from the model object instead of hard-coding them.

```{r}
#| label: a4-konfound
#| include: true

pkonfound(
  est_eff      = 2.5918,
  std_err      = 1.2526,
  n_obs        = 2000,
  n_covariates = 2,
  index        = "RIR"
)

pkonfound(
  est_eff      = 2.5918,
  std_err      = 1.2526,
  n_obs        = 2000,
  n_covariates = 2,
  index        = "IT"
)
```